---
title: "What affects World Happiness?"
author: "Brandon Pugh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    number_sections: FALSE
    theme: spacelab
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
What is happiness? Many have tried to answer this question, and no one really has found a clear, definitive answer to this question. The World Happiness Organization is one of these groups, and have spent years gathering data of different socio-economic factors to compile, what they call, the "life ladder". This is how they measure the happiness of a country, and contains a value from 0 to 10, with 0 being the lowest value of happiness and 10 being the highest. The closer a country's life ladder value, the happier it's citizens are. The purpose of this analysis is to determine what socio-economic factors have an impact on a nation's happiness, and use that information to find out which nations are doing the worse at those significant factors so that the UN can direct policy to try to improve the situation in those countries.

## Data Overview

### Data Source and References
The data that is used in this analysis is contained within the file "HappyData.csv", which is stored in the Happiness folder on the main branch of the GitHub site listed below:

https://github.com/Pughbt/_ISA616HappinessProject 

### Variable Information
The following are the variables that are included in the data set with their definition:

**Country.name** is the name of the country

**year** is the year of which the data was created.

**Life.Ladder** is the happiness score for that country in that specific year. It is on a scale from 0 to 10, with 0 being absolutely unhappy and 10 being extremely happy.

**Log.GDP.per.capita** is the log value of a nation's GDP per capita

**Social.Support** is the percentage of people in a country that says that they have someone to count on in times of trouble.

**Healthy.life.expectancy.at.birth** is the average life expectancy for a person at birth
Freedom.to.make.life.choices is the national average of the citizen's belief on the amount of freedom in their lives

**Generosity** is  the residual of regressing national average of response to the GWP question “Have you donated money to a charity in the past month?” on GDP per capita.

**Perceptions.of.corruption** is the percentage of people in a country that believe that their government has widespred corruption

**Positive.affect** is the average of how many times a person feels happy in a day.

**Negative.affect** is the average of how many times a person feels sad/angry/worried in a day

**Confidence.in.national.government** is the average of how many peole who have confidence in their government

**Democratic.Quality** is how democratic a country's government is. **Delivery.quality** is also measures the same thing.

**Standard.deviation.of.ladder.by.country.year** is an alternative measure of hte inequality of happiness. **Standard.deviation.mean.of.ladder.by.country.year** is the mean of the previous variable

The **GIN**I world index variables measures World bank data for a country

The **most.people.can.be.trusted variables** is the average percentage of people in a country that most people can be trused, with most of the variables being that for different time periods.

### Data Summary

```{r}
##Loading in all required packages for the analysis
pacman:: p_load(DataExplorer, dplyr, corrplot)

## Reading in the data
happiness = read.csv("Happiness/HappyData.csv", stringsAsFactors = TRUE)

##Showing the dimension of the data
dim(happiness)
```

The data has 1848 different observations across 26 variables. Each row is a unique combination of the country name and the year for which data was collected. 

```{r}
## Provides summary statistics for the happiness data set
summary(happiness)
```

Looking at the summary statistics for the variables within the happiness data, we notice several key things. One of which is the year variable. The years that this data was taken was between 2005 and 2019. However, when we look at the actual data set, we find that not every country had their data recorded in each year. Another thing that is clear is the amount of missingness in this data set. Only **Country.Name**, **year**, and **life_ladder** have no missing data in them. Every other variable has some amount of missing data in it, with some of these variables reaching into the thousands of values missing. Another key takeaway is that there are several variables whose values lie between 0 and 1. All of the most.people.can.be.trusted variables, along with social.support, freedom.to.make.life.choices, perceptions.of.corrpution, positive.affect, negative.affect, and confidence.in.national.government, have this characteristic. This means that these variables can be interpreted as the percent of people in these country that feel that way about the factor. For example, if a country had confidence .in.national.government value of .6578, that would mean that 65.78\% of its citizens had trust in the government.

## Data Cleaning and PreProcessing

Before we can start building our model, we first have to go through our data and check various items to see if we can reduce our model. By reducing our model, we can make our analysis that much more efficient and accurate.

### Missing Data
The first thing we should check for is the missingness of our data. The general rule of thumb is that if a variable has more than **50\%** of it's values missing, it is then dropped from the data set. 

```{r}
##Plot the missingness of the data
plot_missing(happiness)
```

As seen in the above chart, the variables that have bars that are either red or purple have their percentage of missing values greater than 50%. These are then dropped from the dataset as they would not provide much information due to their high amounts of missingness.
```{r}
happiness = select(happiness, -Most.people.can.be.trusted..Gallup)
happiness = select(happiness, -Most.people.can.be.trusted..WVS.round.1981.1984)
happiness = select(happiness, -Most.people.can.be.trusted..WVS.round.1989.1993)
happiness = select(happiness, -Most.people.can.be.trusted..WVS.round.1994.1998)
happiness = select(happiness, -Most.people.can.be.trusted..WVS.round.1999.2004)
happiness = select(happiness, -Most.people.can.be.trusted..WVS.round.2005.2009)
happiness = select(happiness, -Most.people.can.be.trusted..WVS.round.2010.2014)
happiness = select(happiness, -GINI.index..World.Bank.estimate.)
```


### Renaming variables

The following variables were renamed due to their length being too long for the visuals that follow in this report to be visible:

**Standard.deviation.of.ladder.by.country.year** was renamed to **sdofhappiness**
**Standard.deviation.Mean.of.ladder.by.country.year** was renamed to **meansdofhappiness**
**GINI.index..World.Bank.estimate...average.2000.2017..unbalanced.panel** was renamed to **GINIWorldBank**
**gini.of.household.income.reported.in.Gallup..by.wp5.year** was renamed to **GINIHousehold**
```{r}
##Rename variables and drop old ones
happiness$sdofhappiness = happiness$Standard.deviation.of.ladder.by.country.year
happiness = select(happiness, -Standard.deviation.of.ladder.by.country.year)
happiness$meansdofhappiness = happiness$Standard.deviation.Mean.of.ladder.by.country.year
happiness = select(happiness, -Standard.deviation.Mean.of.ladder.by.country.year)
happiness$GINIWorldBank = happiness$GINI.index..World.Bank.estimate...average.2000.2017..unbalanced.panel
happiness = select(happiness, -GINI.index..World.Bank.estimate...average.2000.2017..unbalanced.panel)
happiness$GINIHousehold = happiness$gini.of.household.income.reported.in.Gallup..by.wp5.year
happiness = select(happiness, -gini.of.household.income.reported.in.Gallup..by.wp5.year)
```

### Correlation

The next thing that we have to check before we can do our analysis is to check the correlation of the remaining variables in the data set. If we find variables that are highly correlated, one of them will be removed since both of the variables explain the same information. It would make the data redundant in the analysis.

```{r}
##Show the correlation for the variables left in the data set

## Turns the data set into only numeric variables
happiness_numeric = select(happiness, -Country.name)
## Creates a matrix of the correlations 
M = cor(happiness_numeric, use = "complete.obs")

##Creates the correlation plot
corrplot(M, method = "color")
```

In the above correlation plot, we see that **Democratic.Quality** and **Delivery.Quality** are highly correlated with each other. The variable that was dropped was delivery.quality.

```{r}
##Drop highly correlated variabels
happiness = select(happiness, -Delivery.Quality)
```

### Data Aggregation

Now we have to decide how we want to do our analysis. As it stands, we have 1848 observations based on the unique combination of country and year. The way that the data is currently structured is shown in the data table below. If we were to do an analysis of this data as is, we would have 13 year dummy variables with 2005 as the base to show the effect that each year has on happiness. But this does not help us in answering the question of what socio-economical factors affect happiness, as year is not one of those factors. It also doesn't help in providing advice on how to improve happiness, as the other factors can be improved upon by governmental bodis such as the UN. We cannot make any change to the year to improve happiness, as time is outside of human control.

```{r}
head(happiness)
```


Instead of that, I've decided to do an aggregation of the data. This way, we have a data set with each row corresponding to one country, with the other variables being averages for that country across all years that the country was present in the data set.

```{r}
##Aggregating the data
happiness.agg = happiness %>%
  group_by(Country.name) %>%
  summarize(avg.Life.Lader = mean(Life.Ladder, na.rm = T),
            avg.log.GDP.per.capita = mean(Log.GDP.per.capita, na.rm = T),
            avg.social.support = mean(Social.support, na.rm =T),
            avg.life.expectancy = mean(Healthy.life.expectancy.at.birth, na.rm = T),
            avg.freedom = mean(Freedom.to.make.life.choices, na.rm = T),
            avg.generosity = mean(Generosity, na.rm = T),
            avg.perception.of.corrpution = mean(Perceptions.of.corruption, na.rm = T),
            avg.positive.affect = mean(Positive.affect, na.rm = T),
            avg.negative.affect = mean(Negative.affect, na.rm = T),
            avg.confidence.in.government = mean(Confidence.in.national.government, na.rm = T),
            avg.democratic.quality = mean(Democratic.Quality, na.rm = T),
            avg.sd.of.happiness = mean(sdofhappiness, na.rm = T),
            avg.mean.sd.of.happiness = mean(meansdofhappiness, na.rm = T),
            avg.GINI.world.bank = mean(GINIWorldBank, na.rm = T),
            avg.gini.household = mean(GINIHousehold, na.rm = T)
            )
```

Now the data set will look like the table below:

```{r}
head(happiness.agg)
```

This makes the analysis much easier to perform and interpret as we no longer have to worry about time differences in the data.

### Missing Data in the Aggregate

Now that we have aggregated the data, we now have to check to see if there is any missing data with the aggregates.

```{r}
plot_missing(happiness.agg)
```

We see that even after the aggregation, there are still missing values for certain variables. This means that for the certain country, no data was recorded for that variable.

So how do we handle these missing variables? Since there is not a lot of missing values, we can input the data into ourselves. We do this by replacing the missing values with the average value for that variable for the rest of the observations. This way the analysis can be completed without making drastic changes to the data.

```{r}
rows = which(is.nan(happiness.agg$avg.confidence.in.government))
happiness.agg$avg.confidence.in.government[rows] = mean(happiness.agg$avg.confidence.in.government, na.rm = T)

rows1 = which(is.nan(happiness.agg$avg.negative.affect))
happiness.agg$avg.negative.affect[rows1] = mean(happiness.agg$avg.negative.affect, na.rm = T)

rows2 = which(is.nan(happiness.agg$avg.social.support))

happiness.agg$avg.social.support[rows2] = mean(happiness.agg$avg.social.support, na.rm = T)

rows3 = which(is.nan(happiness.agg$avg.democratic.quality))
happiness.agg$avg.democratic.quality[rows3] = mean(happiness.agg$avg.democratic.quality, na.rm = T)

rows4 = which(is.nan(happiness.agg$avg.positive.affect))
happiness.agg$avg.positive.affect[rows4] = mean(happiness.agg$avg.positive.affect, na.rm = T)

rows5 = which(is.nan(happiness.agg$avg.gini.household))
happiness.agg$avg.gini.household[rows5] = mean(happiness.agg$avg.gini.household, na.rm = T)

rows6 = which(is.nan(happiness.agg$avg.log.GDP.per.capita))
happiness.agg$avg.log.GDP.per.capita[rows6] = mean(happiness.agg$avg.log.GDP.per.capita, na.rm = T)

rows7 = which(is.nan(happiness.agg$avg.generosity))
happiness.agg$avg.generosity[rows7] = mean(happiness.agg$avg.generosity, na.rm = T)

rows8 = which(is.nan(happiness.agg$avg.life.expectancy))
happiness.agg$avg.life.expectancy[rows8] = mean(happiness.agg$avg.life.expectancy, na.rm = T)

rows9 = which(is.nan(happiness.agg$avg.perception.of.corrpution))
happiness.agg$avg.perception.of.corrpution[rows9] = mean(happiness.agg$avg.perception.of.corrpution, na.rm= T)

rows10 = which(is.nan(happiness.agg$avg.GINI.world.bank))
happiness.agg$avg.GINI.world.bank[rows10] = mean(happiness.agg$avg.GINI.world.bank, na.rm = T)
```

```{r}
plot_missing(happiness.agg)
```

Now that we have no missing data in this data set, we can start doing our analysis.

## The Model

The first thing we have to do is to determine what type of model we will use to help answer the question of what affects a country's happiness score. Since this question is explanatory in nature, we can use a multiple linear regression model to perform our analysis. We do not include Country.Name in our model as it would not be useful in explaining happiness. We also do not include sdofhappiness and meansdofhappiness variables as they are related to the response variable and would not be useful in explaining happiness. The results of the modle appear below.

```{r}
model1 = lm(avg.Life.Lader~ avg.log.GDP.per.capita + avg.social.support + avg.life.expectancy
            + avg.freedom + avg.generosity + avg.perception.of.corrpution + avg.positive.affect 
            + avg.negative.affect + avg.confidence.in.government + avg.democratic.quality
             + avg.GINI.world.bank +
              avg.gini.household, data = happiness.agg)
summary(model1)
```

### The Results

After the running of the model, we can know see what factors actually affect the happiness of countries around the world. These variables are avg.log.GDP.per.capita, avg.social.support, avg.life.expectancy, avg.freedom, avg.perception.of.corruption, avg.positive.affect, avg.confidence.in.government, and avg.gini.household

For the following variables, an increase in the variable leads to a increase in a country's happiness:

**avg.log.GDP.per.capita**

**avg.social.support**

**avg.life.expectancy**

**avg.freedom**

For the following variables, an increase in the variable leads to a decrease in a country's happiness:

**avg.perception.of.corruption**

**avg.confidence.in.government**

**avg.gini.household**

For almost all of these variables, the way that they affect happiness makes sense. We expect that the richer a country is, the more people that individuals can rely on, the longer a person lives, and the more freedom a person has to live their own life would make a country's populace. We would also expect the more corrupt a government is and the more variability in a household would lead to a country's populace to be less happy. But we wouldn't expect that the higher a country's populace has confidence in their government leads to that populace being unhappy. You would think that the less confident a country is in their leadership, the less happy they would be. But the opposite occurs. Let's take a deeper look at this particular variable

#### The avg.confidence.in.government Variable